Automated assessment of residual plots with computer vision models

### 2024-05-27 paper/paper.tex
wait for review

\begin{abstract}
TBD.
\end{abstract}

@@

In this paper, we develop computer vision models and integrate them into
the residual plots diagnostics workflow, filling the gap of\ldots. The
paper is structured as follows: \ldots{}

@@

\section{Methodology}\label{methodology}

\subsection{Model Specifications}\label{model-specifications}

@@

\subsubsection{Input Formats}\label{input-formats}

@@

\subsubsection{Output Formats}\label{output-formats}

@@

\subsection{Distance Between Residual
Plots}\label{distance-between-residual-plots}

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for assessing residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance metrics
for residual plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

\subsubsection{Residual Distribution}\label{residual-distribution}

@@

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate log-normal distribution, \(P\) will usually be skewed and
has a long tail.

\subsubsection{\texorpdfstring{Distance of \(P\) from
\(Q\)}{Distance of P from Q}}\label{distance-of-p-from-q}

@@

\subsubsection{\texorpdfstring{Non-normal
\(P\)}{Non-normal P}}\label{non-normal-p}

@@

\subsection{Distance Estimation}\label{distance-estimation}

In the previous sections, we have defined a distance measure given in
equation \ref{eq:kl-0} for quantifying the difference between the true
residual distribution \(P\) and an ideal reference distribution \(Q\).
You may have noticed that this distance measure can only be computed
when the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

@@

\subsection{Model Violations Index
(MVI)}\label{sec:model-violations-index}

@@

\subsection{Statistical testing}\label{statistical-testing}

\subsubsection{Lineup Evaluation}\label{lineup-evaluation}

@@

To address this issue while adhering to the principle of visual
inference, we can compare the estimated distance \(\hat{D}\) to the
estimated distances for the null plots in a lineup. Specifically, if a
lineup comprises 20 plots, the null hypothesis \(H_0\) will be rejected
if \(\hat{D}\) exceeds the maximum estimated distance among the
\(m - 1\) null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
participant. The estimated distance serves as a metric to quantify the
difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95\% significance
level can be maintained if the number of plots is a multiple of 20.
Specifically, for lineups comprising \(20t\) plots, where \(l\) is a
positive integer, we reject \(H_0\) if \(\hat{D}\) exceeds 95\%
\({\hat{D}_{null}^{(i)}}\) for \(i = 1, \ldots, 20t-1\). The \(p\)-value
in this case is given by
\(\frac{1}{20t} + \frac{1}{20t}\sum{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)\),
where \(I(\cdot)\) is the indicator function.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the approximated distance. Consequently, quantiles of
the null distribution can be estimated using the sample quantiles
available in statistical software such as R, and these quantiles can be
utilized for decision-making purposes. The details of the sample
quantile computation can be found in \citet{hyndman1996sample}. For
instance, if \(\hat{D}\) is greater than or equal to the 95\% sample
quantile, denoted as \(Q_{null}(0.95)\), we can conclude that the
estimated distance for the true residual plot is significantly different
from the estimated distance for null plots with a 95\% significance
level. Based on our experience, to obtain a stable estimate of the 95\%
quantile, the number of null plots, \(n_{null}\), typically needs to be
at least 100. However, if the null distribution exhibits a long tail, a
larger number of null plots may be required. Alternatively, a
\(p\)-value can be used to represent the probability of observing an
event equally or more extreme than the given event under the null
hypothesis \(H_0\), and it can be estimated by
\(\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

\subsubsection{Bootstrapping}\label{bootstrapping}

@@

\subsection{Data Genearation}\label{data-genearation}

\subsubsection{Simulation Scheme}\label{simulation-scheme}

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data. This approach provided
us with precise label annotations. Additionally, it ensured a large and
diverse training dataset, as we had control over the data generating
process, and the simulation of the training data was relatively
cost-effective.

@@

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
Hermite polynomials \citep{hermite1864nouveau}, the \(\sqrt{(.)}\) and
\((.)^2\) operators are element-wise operators, \(\odot\) is the
Hadamard product, and \(g(., k)\) is a scaling function to enforce the
support of the random vector to be \([-k, k]^n\) defined as

@@

\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength. Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
\centering
\begin{tabular}[t]{ll}
\toprule
Factor & Domain\\
\midrule
j & \{2, 3, ..., 18\}\\
a & {}[-1, 1]\\
b & {}[0, 100]\\
$\beta_1$ & {0, 1}\\
$\beta_2$ & {0, 1}\\
\addlinespace
$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
$\sigma_{X1}$ & {}[0.3, 0.6]\\
\addlinespace
$\sigma_{X2}$ & {}[0.3, 0.6]\\
n & {}[50, 500]\\
\bottomrule
\end{tabular}
\end{table}

@@

Non-normality violations arise from specifying a non-normal distribution
for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
distinct error distributions are considered, including discrete,
uniform, normal, and log-normal distributions, as presented in Figure
\ref{fig:different-e}. Each distribution imparts unique characteristics
to the residuals. The discrete error distribution introduces
discreteness in residuals, while the log-normal distribution typically
yields outliers. Uniform error distribution may result in residuals
filling the entire space of the residual plot. All of these
distributions exhibit visual distinctions from the normal error
distribution.

@@

\subsubsection{Scagnostics}\label{scagnostics}

@@

For each generated residual plot, we computed four scagnostics --
``Monotonic,'' ``Sparse,'' ``Splines,'' and ``Striped'' -- using the
\texttt{cassowaryr} R package \citep{mason2022cassowaryr}. These
computed measures, along with the number of observations from the fitted
model, were provided as the second input for the computer vision model.
While other scagnostics provide valuable insights, they come with high
computational costs and are not suitable for quick inference.

\subsubsection{Balanced Dataset}\label{balanced-dataset}

@@

\subsection{Model Architecture}\label{model-architecture}

@@

\caption{Diagram of the architecture of the optimized computer vision model.}\label{fig:cnn-diag}
\end{figure}

@@

\subsection{Model Training}\label{model-training}

To achieve an optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
Python library \citep{omalley2019kerastuner} for this purpose. A
comprehensive list of hyperparameters is provided in Table
\ref{tab:hyperparameter}.

@@

Our model was trained on the M3 high-performance computing platform,
using TensorFlow and Keras. During training, 80\% of the training data
was utilized for actual training, while the remaining 20\% was used as
validation data. The Bayesian optimization tuner conducted 100 trials to
identify the best hyperparameter values based on validation root mean
square error. The tuner then restored the best epoch of the best model
from the trials. Additionally, we applied early stopping, terminating
the training process if the validation root mean square error fails to
improve for 50 epochs. The maximum allowed epochs was set at 2000,
although no models reached this threshold.

@@

\caption{Hexagonal heatmap for residuals vs predicted distance on training and test data for three optimized models with different input sizes. The red lines are smoothing curves produced by fitting gnealized additive models. The area over the zero line in light pink indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
\end{figure}

@@

\caption{Scatter plots for residuals vs $\sigma$ on test data for the $32 \times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The red lines are smoothing curves produced by fitting gnealized additive models. The area over the zero line in light pink indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
\end{figure}

@@

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate vs the \(\delta\)-difference. In the experiment conducted
in \citet{li2023plot}, participants were allowed to make multiple
selections for a lineup. The weighted detection rate is computed by
assigning weights to each detection. If the participant selects zero
plots, the corresponding weight is 0.05; otherwise, the weight is 1
divided by the number of selections. The \(\delta\)-difference is
defined by \citet{chowdhury2018measuring} as

@@

Figure \ref{fig:delta} indicates that the weighted detection rate
increases as the \(\delta\)-difference increases, particularly when the
\(\delta\)-difference is greater than zero. A negative
\(\delta\)-difference suggests that there is at least one null plot in
the lineup with a stronger visual signal than the true residual plot. In
some instances, the weighted detection rate is close to one, yet the
\(\delta\)-difference is negative. This discrepancy implies that the
distance measure, or the approximated distance, may not perfectly
reflect actual human behaviour.

@@

\caption{A weighted detection rate vs $\delta$-differnence plot.}\label{fig:delta}
\end{figure}

@@

In this section, we will utilize the trained computer vision model on
three example datasets. These include the dataset associated with the
residual plot displaying a ``left-triangle'' shape, as displayed in
Figure \ref{fig:false-finding}, along with the Boston housing dataset,
and the ``dino'' datasets from the \texttt{datasaurus} R package.

@@

The third example presents a situation where the model deviation is
non-typical. Here, the computer vision model and human visual inspection
reject \(H_0\), whereas conventional tests do not. This emphasizes the
benefits of visual inspection and the unique advantage of the computer
vision model, which, like humans, makes decisions based on visual
discoveries.

@@

Figure \ref{fig:false-check} presents the results of the assessment.
Notably, the observed visual signal strength is considerably lower than
the 95\% sample quantile of the null distribution. Moreover, the
bootstrapped distribution suggests that it is highly improbable for the
fitted model to be misspecified as the majority of bootstrapped fitted
models will not be rejected. Thus, for this particular fitted model,
both the visual test and the computer vision model will not reject
\(H_0\). However, the Breusch-Pagan test will reject \(H_0\) because it
can not effectively utilize information from null plots.

@@

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting "left-triangle" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distnace for null plots and bootstrapped plots. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$. (D) PC2 vs PC1 plot for features extracted from the global pooling layer of the computer vision model.  }\label{fig:false-check}
\end{figure}

@@

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the residual plot.
Furthermore, the RESET test yields a very small \(p\)-value. The
approximated distance \(\hat{D}\) significantly exceeds
\(Q_{null}(0.95)\), leading to rejection of \(H_0\). The bootstrapped
distribution also suggests that almost all the bootstrapped fitted
models will be rejected, indicating that the fitted model is unlikely to
be correctly specified.

Additionally, the lineup of this fitted model presented in Figure
\ref{fig:boston-lineup} underscores that the ``U'' shape is a
distinctive characteristic of the true residual plot, absent in any of
the null plots. Consequently, if a visual test is conducted, \(H_0\)
will also be rejected.

@@

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. On the left, the residual plot of the Boston housing fitted model is displayed. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will be rejected since $\hat{D} >= Q_{null}(0.95)$.}\label{fig:boston-check}
\end{figure}

@@

The ``dino'' dataset within the \texttt{datasaurus} R package
exemplifies this scenario. With only two columns, x and y, fitting a
regression model to this data yields a residual plot resembling a
``dinosaur,'' as displayed in Figure \ref{fig:dino-check}.
Unsurprisingly, this distinct residual plot stands out in a lineup, as
shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
would undoubtedly reject \(H_0\).

@@

@@

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. On the left, the residual plot exhitbiting a "dinosaur" shape is displayed. The blue area indicates the distribution of approximated distances for null plots, while the red area represents the distribution of approximated distances for bootstrapped plots. The fitted model will be rejected since $\hat{D} >= Q_{null}(0.95)$.}\label{fig:dino-check}
\end{figure}

@@

\section{Extension to other types of regression
model}\label{extension-to-other-types-of-regression-model}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Use raw residuals.
\end{enumerate}

\begin{itemize}
\tightlist
\item
  D (effect size) for null plots will be greater than zero.
\item
  Can not assume D for actual plots are greater than D for null plots
  any more. -\textgreater{} KL null - normal, KL actual - normal
\item
  Need to use a two-sided test instead of a one-sided test.
\item
  It may fail completely due to the identification issue.
\end{itemize}

>>>>

\begin{abstract}
Plotting residuals is a standard practice in linear regression
diagnostics, essential for identifying deviations from model assumptions
such as linearity, homoscedasticity, and normality. Visual inference
provides an inferential framework to assess whether residual plots
contain patterns inconsistent with model assumptions, typically using a
lineup protocol. However, the lineup protocol's reliance on human
judgment limits its scalability. This study addresses this limitation by
automating the interpretation of residual plots using computer vision
models. We develop a distance measure based on Kullback-Leibler
divergence to quantify the disparity between the residual distribution
of a fitted classical normal linear regression model and the reference
distribution. We propose a computer vision model to estimate this
distance from residual plots, facilitating formal statistical testing
and bootstrapping techniques to assess model specification. Our computer
vision model shows strong performance, though it performs slightly less
effectively on non-linearity visual patterns. The statistical tests
based on the estimated distance exhibit lower sensitivity than
conventional tests but higher sensitivity than human visual tests.
Examples demonstrate the method's effectiveness across different
scenarios, highlighting its value in automating the diagnostic process
and supplementing traditional methods.
\end{abstract}

@@

In this paper, we develop computer vision models and integrate them into
the residual plots diagnostics workflow, addressing the need for
automated visual inference. The paper is structured as follows. Section
\ref{model-specifications} discusses various specifications of the
computer vision models. Section \ref{distance-between-residual-plots}
defines the distance measure used to measure model violations. Section
\ref{distance-estimation} explains how the computer vision models
estimate this distance measure. Section \ref{statistical-testing} covers
the statistical testing based on the estimated distance. Sections
\ref{data-generation}, \ref{model-architecture}, and
\ref{model-training} detail the data preparation, model architecture,
and training process, respectively. The results are presented in Section
\ref{results}. Finally, we conclude with a discussion of our findings
and propose ideas for future research directions.

@@

\section{Model Specifications}\label{model-specifications}

@@

\subsection{Input Formats}\label{input-formats}

@@

\subsection{Output Formats}\label{output-formats}

@@

\section{Distance Between Residual
Plots}\label{distance-between-residual-plots}

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for assessing residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance measures
for residual plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

\subsection{Residual Distribution}\label{residual-distribution}

@@

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but missing from \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted regression model
will not be the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

\subsection{\texorpdfstring{Distance of \(P\) from
\(Q\)}{Distance of P from Q}}\label{distance-of-p-from-q}

@@

\subsection{\texorpdfstring{Non-normal
\(P\)}{Non-normal P}}\label{non-normal-p}

@@

\section{Distance Estimation}\label{distance-estimation}

In the previous sections, we have defined a distance measure given in
equation \ref{eq:kl-0} for quantifying the difference between the true
residual distribution \(P\) and an ideal reference distribution \(Q\).
It can be noticed that this distance measure can only be computed when
the data generating process is known. In reality, we often have no
knowledge about the data generating process, otherwise we do not need to
fit a linear regression model in the first place.

@@

\section{Model Violations Index (MVI)}\label{sec:model-violations-index}

@@

\section{Statistical testing}\label{statistical-testing}

\subsection{Lineup Evaluation}\label{lineup-evaluation}

@@

To address this issue while adhering to the principle of visual
inference, we can compare the estimated distance \(\hat{D}\) to the
estimated distances for the null plots in a lineup. Specifically, if a
lineup comprises 20 plots, the null hypothesis \(H_0\) will be rejected
if \(\hat{D}\) exceeds the maximum estimated distance among the
\(m - 1\) null plots, denoted as
\(\max\limits_{1 \leq i \leq m-1} {\hat{D}_{null}^{(i)}}\), where
\(\hat{D}_{null}^{(i)}\) represents the estimated distance for the
\(i\)-th null plot. This approach is equivalent to the typical lineup
protocol requiring a 95\% significance level, where \(H_0\) is rejected
if the data plot is identified as the most distinct plot by the sole
observer. The estimated distance serves as a metric to quantify the
difference between the data plot and the null plots, as intended.

For lineups consisting of more than 20 plots, the 95\% significance
level can be maintained if the number of plots is a multiple of 20.
Specifically, for lineups comprising \(20t\) plots, where \(t\) is a
positive integer, we reject \(H_0\) if \(\hat{D}\) exceeds 95\%
\({\hat{D}_{null}^{(i)}}\) for \(i = 1, \ldots, 20t-1\). The \(p\)-value
in this case is given by
\(\frac{1}{20t} + \frac{1}{20t}\sum_{i=1}^{20t-1} I\left(\hat{D}_{null}^{(i)} > \hat{D}\right)\),
where \(I(\cdot)\) is the indicator function.

Moreover, if the number of plots in a lineup, denoted by \(m\), is
sufficiently large, the empirical distribution of
\({\hat{D}_{null}^{(i)}}\) can be viewed as an approximation of the null
distribution of the estimated distance. Consequently, quantiles of the
null distribution can be estimated using the sample quantiles available
in statistical software such as R, and these quantiles can be utilized
for decision-making purposes. The details of the sample quantile
computation can be found in \citet{hyndman1996sample}. For instance, if
\(\hat{D}\) is greater than or equal to the 95\% sample quantile,
denoted as \(Q_{null}(0.95)\), we can conclude that the estimated
distance for the true residual plot is significantly different from the
estimated distance for null plots with a 95\% significance level. Based
on our experience, to obtain a stable estimate of the 95\% quantile, the
number of null plots, \(n_{null}\), typically needs to be at least 100.
However, if the null distribution exhibits a long tail, a larger number
of null plots may be required. Alternatively, a \(p\)-value can be used
to represent the probability of observing an event equally or more
extreme than the given event under the null hypothesis \(H_0\), and it
can be estimated by
\(\frac{1}{m}\sum_{i=1}^{m-1}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

\subsection{Bootstrapping}\label{bootstrapping}

@@

\section{Data Generation}\label{data-generation}

\subsection{Simulation Scheme}\label{simulation-scheme}

While observational data is frequently employed in training models for
real-world applications, the data generating process of observational
data often remains unknown, making computation for our target variable
\(D\) unattainable. Consequently, the computer vision models developed
in this study were trained using synthetic data, including 80000
training images and 8000 test images. This approach provided us with
precise label annotations. Additionally, it ensured a large and diverse
training dataset, as we had control over the data generating process,
and the simulation of the training data was relatively cost-effective.

@@

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
Hermite polynomials \citep{hermite1864nouveau}, the \(\sqrt{(.)}\) and
\((.)^2\) operators are element-wise operators, \(\odot\) is the
Hadamard product, and \(g(\boldsymbol{x}, k)\) is a scaling function to
enforce the support of the random vector to be \([-k, k]^n\) defined as

@@

\caption{\label{tab:factor}Factors used in the data generating process for synthetic data simulation. Factor $j$ and $a$ controls the non-linearity shape and the heteroskedasticity shape respectively. Factor $b$, $\sigma_\varepsilon$ and $n$ control the signal strength. Factor $\text{dist}_\varepsilon$, $\text{dist}_{x1}$ and $\text{dist}_{x2}$ specifies the distribution of $\varepsilon$, $X_1$ and $X_2$ respectively.}
\centering
\begin{tabular}[t]{ll}
\toprule
Factor & Domain\\
\midrule
j & \{2, 3, ..., 18\}\\
a & {}[-1, 1]\\
b & {}[0, 100]\\
$\beta_1$ & \{0, 1\}\\
$\beta_2$ & \{0, 1\}\\
\addlinespace
$\text{dist}_\varepsilon$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x1}$ & \{discrete, uniform, normal, lognormal\}\\
$\text{dist}_{x2}$ & \{discrete, uniform, normal, lognormal\}\\
$\sigma_{\varepsilon}$ & {}[0.0625, 9]\\
$\sigma_{X1}$ & {}[0.3, 0.6]\\
\addlinespace
$\sigma_{X2}$ & {}[0.3, 0.6]\\
n & {}[50, 500]\\
\bottomrule
\end{tabular}
\end{table}

@@

Non-normality violations arise from specifying a non-normal distribution
for \(\boldsymbol{\varepsilon}\). In the synthetic data simulation, four
distinct error distributions are considered, including discrete,
uniform, normal, and lognormal distributions, as presented in Figure
\ref{fig:different-e}. Each distribution imparts unique characteristics
to the residuals. The discrete error distribution introduces
discreteness in residuals, while the lognormal distribution typically
yields outliers. Uniform error distribution may result in residuals
filling the entire space of the residual plot. All of these
distributions exhibit visual distinctions from the normal error
distribution.

@@

\subsection{Scagnostics}\label{scagnostics}

@@

For each generated residual plot, we computed four scagnostics --
``Monotonic'', ``Sparse'', ``Splines'', and ``Striped'' -- using the
\texttt{cassowaryr} R package \citep{mason2022cassowaryr}. These
computed measures, along with the number of observations from the fitted
model, were provided as the second input for the computer vision model.
While other scagnostics provide valuable insights, they come with high
computational costs and are not suitable for quick inference.

\subsection{Balanced Dataset}\label{balanced-dataset}

@@

\section{Model Architecture}\label{model-architecture}

@@

\caption{Diagram of the architecture of the optimized computer vision model. Numbers at the bottom of each box show the shape of the output of each layer. The band of each box drawn in a darker colour indicates the use of the rectified linear unit activation function.  Yellow boxes are 2D convolutional layers, orange boxes are pooling layers, the grey box is the concatenation layer, and the purple boxes are dense layers.}\label{fig:cnn-diag}
\end{figure}

@@

\section{Model Training}\label{model-training}

To achieve a near-optimal deep learning model, hyperparameters like the
learning rate often need to be fine-tuned using a tuner. In our study,
we utilized the Bayesian optimization tuner from the \texttt{KerasTuner}
Python library \citep{omalley2019kerastuner} for this purpose. A
comprehensive list of hyperparameters is provided in Table
\ref{tab:hyperparameter}.

@@

Our model was trained on the MASSIVE M3 high-performance computing
platform \citep{goscinski2014multi}, using TensorFlow
\citep{abadi2016tensorflow} and Keras \citep{chollet2015keras}. During
training, 80\% of the training data was utilized for actual training,
while the remaining 20\% was used as validation data. The Bayesian
optimization tuner conducted 100 trials to identify the best
hyperparameter values based on validation root mean square error. The
tuner then restored the best epoch of the best model from the trials.
Additionally, we applied early stopping, terminating the training
process if the validation root mean square error fails to improve for 50
epochs. The maximum allowed epochs was set at 2000, although no models
reached this threshold.

@@

\caption{Hexagonal heatmap for residuals vs predicted distance on training and test data for three optimized models with different input sizes. The brown lines are smoothing curves produced by fitting gnealized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:model-performance}
\end{figure}

@@

\caption{Scatter plots for residuals vs $\sigma$ on test data for the $32 \times 32$ optimized model. The data is grouped by whether the regression has only non-linearity violation, and whether it includes a second predictor in the regression formula. The brown lines are smoothing curves produced by fitting gnealized additive models. The area over the zero line in light yellow indicates under-prediction, and the area under the zero line in light green indicates over-prediction.}\label{fig:over-under}
\end{figure}

@@

In the experiment conducted in \citet{li2023plot}, participants were
allowed to make multiple selections for a lineup. The weighted detection
rate is computed by assigning weights to each detection. If the
participant selects zero plots, the corresponding weight is 0.05;
otherwise, the weight is 1 divided by the number of selections. The
\(\delta\)-difference is defined by \citet{chowdhury2018measuring} as

@@

Figure \ref{fig:delta} displays the scatter plot of the weighted
detection rate vs the \(\delta\)-difference. It indicates that the
weighted detection rate increases as the \(\delta\)-difference
increases, particularly when the \(\delta\)-difference is greater than
zero. A negative \(\delta\)-difference suggests that there is at least
one null plot in the lineup with a stronger visual signal than the true
residual plot. In some instances, the weighted detection rate is close
to one, yet the \(\delta\)-difference is negative. This discrepancy
implies that the distance measure, or the estimated distance, may not
perfectly reflect actual human behaviour.

@@

\caption{A weighted detection rate vs $\delta$-differnence plot. The brown line is smoothing curve produced by fitting gnealized additive models}\label{fig:delta}
\end{figure}

@@

In this section, we will present the performance of trained computer
vision model on three example datasets. These include the dataset
associated with the residual plot displaying a ``left-triangle'' shape,
as displayed in Figure \ref{fig:false-finding}, along with the Boston
housing dataset, and the ``dino'' datasets from the \texttt{datasaurus}
R package.

@@

The third example presents a situation where the model deviation is
non-typical. Here, the computer vision model and human visual inspection
reject \(H_0\), whereas some commonly used conventional tests do not.
This emphasizes the benefits of visual inspection and the unique
advantage of the computer vision model, which, like humans, makes
decisions based on visual discoveries.

@@

Figure \ref{fig:false-check} presents the results of the assessment by
the computer vision model. Notably, the observed visual signal strength
is considerably lower than the 95\% sample quantile of the null
distribution. Moreover, the bootstrapped distribution suggests that it
is highly improbable for the fitted model to be misspecified as the
majority of bootstrapped fitted models will not be rejected. Thus, for
this particular fitted model, both the visual test and the computer
vision model will not reject \(H_0\). However, the Breusch-Pagan test
will reject \(H_0\) because it can not effectively utilize information
from null plots.

@@

\caption{A summary of the residual plot assessment evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "left-triangle" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distnace for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will not be rejected since $\hat{D} < Q_{null}(0.95)$. (D) PC2 vs PC1 plot for features extracted from the global pooling layer of the computer vision model.  }\label{fig:false-check}
\end{figure}

@@

Figure \ref{fig:boston-check} displays the residual plot and the
assessment conducted by the computer vision model. A clear non-linearity
pattern resembling a ``U'' shape is shown in the plot A. Furthermore,
the RESET test yields a very small \(p\)-value. The estimated distance
\(\hat{D}\) significantly exceeds \(Q_{null}(0.95)\), leading to
rejection of \(H_0\). The bootstrapped distribution also suggests that
almost all the bootstrapped fitted models will be rejected, indicating
that the fitted model is unlikely to be correctly specified. The
attention map in plot B suggests the center of the image has higher
leverage than other areas, and it is the turning point of the ``U''
shape. The principal component analysis provided in plot D shows two
distinct clusters of data points, further underlines the visual
differences between bootstrapped plots and null plots. This coincides
the findings from Figure \ref{fig:boston-lineup}, where the true plot
exhibiting a ``U'' shape is visually distinctive from null plots. If a
visual test is conducted by human, \(H_0\) will also be rejected.

@@

\caption{A summary of the residual plot assessment for the Boston housing fitted model evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "U" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distnace for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$. (D) PC2 vs PC1 plot for features extracted from the global pooling layer of the computer vision model. }\label{fig:boston-check}
\end{figure}

@@

The ``dino'' dataset within the \texttt{datasaurus} R package
exemplifies this scenario. With only two columns, x and y, fitting a
regression model to this data yields a residual plot resembling a
``dinosaur,'' as displayed in plot A of Figure \ref{fig:dino-check}.
Unsurprisingly, this distinct residual plot stands out in a lineup, as
shown in Figure \ref{fig:dino-lineup}. A visual test conducted by humans
would undoubtedly reject \(H_0\).

@@

More importantly, the attention map in plot B of Figure
\ref{fig:dino-check} clearly exhibits a ``dinosaur'' shape, strongly
suggesting the prediction of the distance is based on human perceptible
visual patterns. The computer vision model is also capable of capturing
the contour or the outline of the embedded shape, just like human being
reading residual plots. The principle component analysis in plot D of
Figure \ref{fig:dino-check} also shows that the cluster of bootstrapped
plots is at the corner of the cluster of null plots.

More importantly, the attention map in plot B of Figure
\ref{fig:dino-check} clearly exhibits a ``dinosaur'' shape, strongly
suggesting that the distance prediction is based on human-perceptible
visual patterns. The computer vision model effectively captures the
contour or outline of the embedded shape, similar to how humans
interpret residual plots. Additionally, the principal component analysis
in plot D of Figure \ref{fig:dino-check} demonstrates that the cluster
of bootstrapped plots is positioned at the corner of the cluster of null
plots.

@@

\caption{A summary of the residual plot assessment for the Datasaurus fitted model evaluted on 200 null plots and 200 bootstrapped plots. (A) The true residual plot exhibiting a "dinosaur" shape. (B) The attention map produced by computing the gradient of the output with respect to the greyscale input.  (C) The density plot of estimated distnace for null plots and bootstrapped plots. The blue area indicates the distribution of estimated distances for bootstrapped plots, while the yellow area represents the distribution of estimated distances for null plots. The fitted model will be rejected since $\hat{D} \geq Q_{null}(0.95)$. (D) PC2 vs PC1 plot for features extracted from the global pooling layer of the computer vision model.}\label{fig:dino-check}
\end{figure}

@@

\section{Limitations and Future Work}\label{limitations-and-future-work}

Despite the computer vision model performing well with general cases
under the synthetic data generation scheme and the three examples used
in this paper, this study has several limitations that could guide
future work.
